---
title: "Correlation (Instructional Worksheet)"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<script src="hideOutput.js"></script>

<link rel="stylesheet" type="text/css" href="fold.css">



#\textcolor{blue}{1.1} Scatterplot

A scatterplot can be used to visually compare two variables for the same subject. In R, a scatterplot can be made using the *plot()* function that we learned in Module 5. In fact we have actually already made scatterplots before!

Let's use the *iris* built-in dataset that we have previously used, and compare the petal length and width visually using a scatterplot.

```{r eval = FALSE}
plot(iris$Petal.Length, iris$Petal.Width, xlab = "Petal Length", ylab = "Petal Width")
```
Looking at this plot, is there a positive or negative relationship between petal length and width? Is the relationship linear or non-linear?


#\textcolor{blue}{1.2} Correlation Coefficient (Pearson's *r*)
We can calculate Pearson's *r* correlation coefficient using the *cor.test()* function in R. 
The default version of this function is to run the Pearson's correlation. It is also possible to calculate other correlation coefficients using this function.

```{r eval = FALSE}
cor.test(iris$Petal.Length, iris$Petal.Width)
```
Looking at the results, the correlation coefficient is 0.96. This value is positive, so as petal length increases so does petal width, and very close to 1. The variables are highly correlated, which we know because the p-value is < 0.05.

We are basing these results on the $\alpha = 0.05$ significance level, but we can also change if we are instead interested in $\alpha = 0.01$.
```{r eval = FALSE}
cor.test(iris$Petal.Length, iris$Petal.Width, conf.level = 0.99)
```



#\textcolor{blue}{1.3} Problem Set

1. Using the *trees* built-in dataset, which gives the girth, height, and volume of black cherry trees, create a scatterplot of height versus volume. Is there a positive or negative relationship?

2. Calculate the Pearson's *r* correlation coefficient. What is the coefficient? Are the variables significantly correlated? How do you know?

3. What is the relationship between girth and volume for black cherry trees. Create a scatterplot and report the correlation coefficient.

4. Which variable is most correlated with volume?

After you have completed the problem set, click the Code button to see the solution.


<div class="fold s">

```{r eval = FALSE}
plot(trees$Height, trees$Volume, xlab = "Height", ylab = "Volume", main = "Black Cherry Trees Height Vs. Volume")
cor.test(trees$Height, trees$Volume)
#r = 0.598
#significant (p-value 0.0003784)
plot(trees$Girth, trees$Volume, xlab = "Girth", ylab = "Volume", main = "BlackC Cherry Trees Girth Vs. Volume")
cor.test(trees$Girth, trees$Volume)
#r = 0.967
#significant (p-value <2.2e-16)

#Girth is most correlated with volume (0.967 vs. 0.598)

```

</div>

