---
title: "Correlation (Instructional Worksheet)"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<script src="hideOutput.js"></script>

<link rel="stylesheet" type="text/css" href="fold.css">



#\textcolor{blue}{1.1} Scatterplot

A scatterplot can be used to visually compare two variables for the same subject. In R, a scatterplot can be made using the *plot()* function that we learned in Module 5. In fact we have actually already made scatterplots before!

Let's use the *women* built-in dataset that we have previously used, and compare the women's height and weight visually using a scatterplot.

```{r eval = FALSE}
plot(women$height, women$weight, xlab = "Height", ylab = "Weight")
```
Looking at this plot, is there a positive or negative relationship between women's height and weight? Is the relationship linear or non-linear?


#\textcolor{blue}{1.2} Correlation Coefficient (Pearson's *r*)
We can calculate Pearson's *r* correlation coefficient using the *cor.test()* function in R. 
The default version of this function is to run the Pearson's correlation. It is also possible to calculate other correlation coefficients using this function.

```{r eval = FALSE}
cor.test(women$height, women$weight)
```
Looking at the results, the correlation coefficient is 0.995. This value is positive, so as height increases so does weight, and very close to 1. The variables are highly correlated, which we know because the p-value is < 0.05.

We are basing these results on the $\alpha = 0.05$ significance level, but we can also change if we are instead interested in $\alpha = 0.01$.
```{r eval = FALSE}
cor.test(women$height, women$weight, conf.level = 0.99)
```



#\textcolor{blue}{1.3} Problem Set

1. Using the *leuk* built-in dataset (in package *MASS*), which gives the white blood count and survival time for patients with leukemia, create a scatterplot of white blood count vs. survival time. Is there a positive or negative relationship?

2. Calculate the Pearson's *r* correlation coefficient. What is the coefficient? Are the variables significantly correlated? How do you know?


After you have completed the problem set, click the Code button to see the solution.


<div class="fold s">

```{r eval = FALSE}
plot(leuk$wbc, leuk$time, xlab = "White Blood Count", ylab = "Survival Time (weeks)", main = "Leukemia Patients White Blood Count vs. Survival Time")
cor.test(leuk$wbc, leuk$time)
#r = -0.329
#not significant (p-value 0.06117)
```

</div>

