---
title: "Dispersion in R (Instructional Worksheet)"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<script src="hideOutput.js"></script>

<link rel="stylesheet" type="text/css" href="fold.css">

#\textcolor{blue}{1.1} Measuring Dispersion

We will continue using the *leuk* dataset from the last problem set to explore measures of dispersion, the spread or variability in a set of data. We are interested in the dispersion measures for white blood count. 

Range: difference between the minimum and maximum values

```{r eval = FALSE}
range(leuk$wbc)
```

Variance: measure of the distance of each value from the mean value 

```{r eval = FALSE}
var(leuk$wbc)
```

Standard deviation: the square root of variance (how much the values typically vary from the average value)

```{r eval = FALSE}
sd(leuk$wbc)
```

\newpage

#\textcolor{blue}{1.2} Problem Set

1. We will continue with the the *leuk* dataset. What is the range of values for survival time in weeks?  

2. What is the variance and standard deviation of survival time in weeks?

3. What is the variance and standard deviation of survival time for patients who are AG positive? Is it higher or lower than the overall survival time for patients? (Hint: look at the help file (help(leuk)) to figure out which patients are AG positive)


After you have completed the problem set, click the Code button to see the solution.


<div class="fold s">

```{r eval = FALSE}
range(leuk$time)

var(leuk$time)
sd(leuk$time)

var(leuk$time[leuk$ag == "present"])
sd(leuk$time[leuk$ag == "present"])
#var and sd are higher for patients who are AG positive than overall
```

</div>