---
title: "Non-Parametric Statistics: Chi-Square (Instructional Worksheet)"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<script src="hideOutput.js"></script>

<link rel="stylesheet" type="text/css" href="fold.css">



#\textcolor{blue}{1.1} One Variable Chi-Square: Goodness of Fit

Goal: Compare categorical data with a theoretical distribution
We want to test the following 2 hypotheses:
(1) data follows the theoretical distribution
(2) data does not follow the theoretical distribution

Suppose we have a clinical trial, out of the 929 patients participating in the trial there are 705 who benefit from the treatment and 224 who see no effect of the trial. The expected proportion is 3/4 will benefit from the trail and 1/4 will see no effects. Let's run a one variable chi-square test to see if the patients in the clinical trial follow the expected proportions or not. We will use the *chisq.test*()* function. 

```{r eval = FALSE}
trial <- data.frame(outcome = c("benefit", "noeffect"), freq = c(705, 224))
chisq.test(trial$freq, p = c(0.75, 0.25))
```
In the output, we have the Chi-square statitics, the degrees of freedom, and the p-value. Based on these results, we fail to reject the null hypothesis because the p-value is greater than 0.05. This means that the data follows the theoretical distribution (i.e., the data follows the expected proportions).


#\textcolor{blue}{1.2} Two Variable Chi-Square: Test of Independence

Goal: Determine whether or not two or more variables are related (i.e., not independent)
We want to test the following 2 hypotheses:
(1) variables are independent
(2) variables are not independent

Using our clinical trial example from above, suppose that in addition to the outcome of the trial, we also know whether or not the patient survived to the following year. We are interested in whether or not survival is related to the outcome from the trial. Therefore, our null hypothesis is that there is no relationship between trial outcome and survival, and our alternative hypothesis is that there is a relationship between trial outcome and survival. 

```{r eval = FALSE}
trial$surv <- c(448, 103)
tab <- xtabs(cbind(freq, surv) ~ outcome, trial)
summary(tab)
```

Looking at the results, we again are given the Chi-squared statistic (Chisq), the degrees of freedom (df), and the p-value. With a p-value of 0.01514, we reject our null hypothesis (i.e., support the alternative hypothesis). We can conclude that there is a relationship between trial outcome and survival.

#\textcolor{blue}{1.3} Problem Set

1. You are studying eye color in a class of students. You complete a survey of the current class and discover there are 27 students with blue eyes and 43 students with brown eyes. Last year you did the same survey in your class and there were 16 students with blue eyes and 44 students with brown eyes. Does the current year students match the expected distribution based on the class from the year before, or is there a difference between the observed and expected values? How do you know? (Hint: your expected values have to be probabilities - so make sure you take the number of students for each eye color divided by the total number of students surveyed)

2. Suppose that during our survey, we also collected data on the dominant writing hand for each student. Of the blue eyed students, there were 25 who were right-hand dominant, and out of the brown-eyed students there were 36 who were right-hand dominant. Is the number of students who were blue-eyed vs. brown-eyed dependent on dominant writing hand (i.e., are the two variables independent)? How do you know? 


After you have completed the problem set, click the Code button to see the solution.


<div class="fold s">

```{r eval = FALSE}
student <- data.frame(student = c("blue", "brown"), freq = c(27, 43))
chisq.test(student$freq, p = c(16/60, 44/60))
#encounter rates in current survey are not the same as expected encounter rates from year before
#p-value 0.0243 - reject null hypothesis

student$rh <- c(25, 36)
tab <- xtabs(cbind(freq,rh) ~ student, student)
summary(tab)
#p-value 0.7783
#the two variables are independent - fail to reject null hypothesis
#the eye color of students was not dependent on the dominant writing hand
```

</div>

