---
title: "Standard Scores (Instructional Worksheet)"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<script src="hideOutput.js"></script>

<link rel="stylesheet" type="text/css" href="fold.css">


#\textcolor{blue}{1.1} Normal Distribution

Using R we can generate random numbers within a distribution, such as the normal distribution. This can be useful if we want so simulate data.

For a normal distribution, the standard normal distribution have a mean of 0 and a standard deviation of 1. If we want to generate 100 numbers using the standard normal distribution we do the following:

```{r eval = FALSE}
rnorm(100)
```

We can also generate numbers within a normal distribution with a given mean or standard deviation. For example, assume the mean is 3 and the standard deviation is 1.5. This time we want to generate 150 numbers. 

```{r eval = FALSE}
rnorm(150, mean = 3, sd = 1.5)
```

We can graph the results to show that they follow the normal distribution using a histogram.

```{r eval = FALSE}
r <- rnorm(150, mean = 3, sd = 1.5)
hist(r)
```

This graph might not look exactly like the normal distribution since we have a small sample size, but try it again generating 1000 numbers and look at the plot. Does this look more as you expected?

>###Note: You can add a vertical line at the mean to make sure it is distributed around your given mean value. Make this line blue!

```{r eval = FALSE}
r = rnorm(1000, mean = 3, sd = 1.5)
hist(r)
abline(v=3, col="blue")
```


#\textcolor{blue}{1.2} Calculating Z-scores

A z-score, which is a type of standard score, is the number of standard deviations a data point is from the mean.
The z-score is calculated using basic subtraction and division, and is very straightforward in R.

Let's assume the mean survival time for patients in a clinical trial is 500 days. The standard deviation of the survival time is 30 days.

In a new trial, the mean survival time is 525 days. What is the z-score?

```{r eval = FALSE}
z = (525 - 500)/30
```

Now, assume the mean survival time is 300 days in the next trial. What is the z-score?

```{r eval = FALSE}
z = (300-500)/30
```

>###Note: The parantheses are necessary in this equation to get the correct z-score - what happens if you remove them?


\newpage

#\textcolor{blue}{1.3} Problem Set

1. Generate 1500 random numbers under the standard normal distribution. What is the mean and standard deviation of the standard normal distribution?

2. Generate 10,000 random numbers with a mean of 5 and a standard deviation of 1. 

3. Plot the generated data from question 2 as a histogram and add a vertical line at the mean in blue.

4. The mean white blood count for a group of patients is 30000. The standard deviation of counts is 10000. What is the z-score for a patient that has a white blood count of 52000?

5. What is the z-score for a patient that has a white blood count of 9400?


After you have completed the problem set, click the Code button to see the solution.


<div class="fold s">

```{r eval = FALSE}
rnorm(1500)
#mean = 0, sd = 1

rnorm(10000, mean = 5, sd = 1)

r = rnorm(10000, mean = 5, sd = 1)
hist(r)
abline(v = 5, col = "blue")

z = (52000-30000)/10000

z = (9400-30000)/10000

```

</div>